Nowcasting Covid-19 statistics reported with delay: a case-study of Sweden

### 2020-06-11 output/paper/appendix.tex
tex

\subsection{Priors}
For the $\mv{\alpha}$ and $\mv{\beta}$ parameters we use an (improper) uniform prior. For the deaths, $\mv{d}$, one could imagine several different prior ideally some sort of epidemiological model. However, here we just assume a log-Gaussian Cox processes \cite{Moller1998_log_gaussian}, but instead of Poisson distribution we use a negative binomial to handle possible over dispersion. The latent Gaussian processes has a intrinsic random walk distribution \cite{Rue2005_gaussian_markov} i.e.
\begin{align*}
\log(\mu_i) - \log(\mu_{i-1}) &\sim N(0,\sigma^2),\\
d_i| \lambda_i  &\sim NegBin(\mu_i, \phi).
\end{align*}
This model is created to create a temporal smoothing between the reported deaths.
For the hyperparameter $\sigma^2$ we impose a inverse Gamma distribution, this prior is suitable here because it guarantees that the process is not constant ($\sigma^2=0$) which we know is not the case.
\subsection{Full model}
Putting the likelihood and priors together we get the following hierarchical Bayesian model
\begin{align*}
\sigma^2 &\sim \Gamma(1,0.01) \\
\phi &\sim \Gamma(1,0.01) \\
\alpha_k &\sim U[0,\infty] \\
\beta_k &\sim U[0,\infty] \\
\alpha_k^H &\sim U[0,\infty] \\
\beta_k^H &\sim U[0,\infty] \\
\log(\mu_i) - \log(\mu{i-1}) &\sim N(0,\sigma^2)\\
d_i| \lambda_i  &\sim NegBin(\mu_i,\phi) \\
p_{i,j}|  \mv{\alpha},\mv{\beta}, \mv{\alpha}^H,\mv{\beta}^H &\sim Beta(\alpha^H_j\alpha_{min(j-i,k)},\beta^H_j\beta_{min(j-i,k)}) \\
r_{i,j}|d_i,\mv{r}_{1,1:j},p &\sim \pi Bin(d_i - \sum_{k=1}^{j-1}r_{i,k},p_0)+ (1-\pi) Bin(d_i - \sum_{k=1}^{j-1}r_{i,k}, p_{i,j}),
\end{align*}
where where and $j\leq i$ and $i=1,\ldots,T$.

\section{Inference}
As the main goal to generate inference of the number of death $\mv{d}$ is through the posterior distribution of number of deaths $\mv{d}$ given the observations $\mv{r}$.
In order to generate samples from this distribution we use a Markov Chain Monte Carlo method \cite{Brooks2011_handbook_markov}. In more detail we use a blocked Gibbs sampler, which generates samples in the following sequence:
\begin{itemize}
	\item  We sample $\mv{\alpha},\mv{\beta}, \mv{\alpha}^H,\mv{\beta}^H|\mv{d}, \mv{r}$ using the fact that one can integrate out $p$ in the model, and then  $\mv{d}|\mv{\alpha},\mv{\beta}, \mv{\alpha}^H,\mv{\beta}^H,\mv{r},\mv{\lambda}$  follows a Beta-Binomial distribution. Here to we use an adaptive MALA (\cite{Atchade2006_adaptive_version}) to sample from these parameters.
	\item  To sample $\mv{d}|\mv{\alpha},\mv{\beta}, \mv{\alpha}^H,\mv{\beta}^H,\mv{r},\mv{\lambda}$, that each death, $d_i$ is conditionally independent, and we just use a Metropolis Hastings random walk to sample each one.
	\item To sample $\mv{\lambda} | \mv{d},\sigma^2$ we again use an adaptive MALA.
	\item Finally We sample $\sigma^2|\mv{d}$,and  $p_0,\pi$ directly since this distribution is explicit, and $\phi$ using a MH-RW.
\end{itemize}

>>>>

\subsection{Priors}
For the $\mv{\alpha}$ and $\mv{\beta}$ parameters we use an (improper) uniform prior. For the deaths, $\mv{d}$, one could imagine several different prior ideally some sort of epidemiological model. However, here we just assume a log-Gaussian Cox processes \citep{Moller1998_log_gaussian}, but instead of Poisson distribution we use a negative binomial to handle possible over dispersion. The latent Gaussian processes has a intrinsic random walk distribution \citep{Rue2005_gaussian_markov} i.e.
\begin{align*}
\log(\mu_i) - \log(\mu_{i-1}) &\sim N(0,\sigma^2),\\
d_i| \lambda_i  &\sim NegBin(\mu_i, \phi).
\end{align*}
This model is created to create a temporal smoothing between the reported deaths.
For the hyperparameter $\sigma^2$ we impose a inverse Gamma distribution, this prior is suitable here because it guarantees that the process is not constant ($\sigma^2=0$) which we know is not the case.
\subsection{Full model}
Putting the likelihood and priors together we get the following hierarchical Bayesian model
\begin{align*}
\sigma^2 &\sim \Gamma(1,0.01) \\
\phi &\sim \Gamma(1,0.01) \\
\alpha_k &\sim U[0,\infty] \\
\beta_k &\sim U[0,\infty] \\
\alpha_k^H &\sim U[0,\infty] \\
\beta_k^H &\sim U[0,\infty] \\
\log(\mu_i) - \log(\mu{i-1}) &\sim N(0,\sigma^2)\\
d_i| \lambda_i  &\sim NegBin(\mu_i,\phi) \\
p_{i,j}|  \mv{\alpha},\mv{\beta}, \mv{\alpha}^H,\mv{\beta}^H &\sim Beta(\alpha^H_j\alpha_{min(j-i,k)},\beta^H_j\beta_{min(j-i,k)}) \\
r_{i,j}|d_i,\mv{r}_{1,1:j},p &\sim \pi Bin(d_i - \sum_{k=1}^{j-1}r_{i,k},p_0)+ (1-\pi) Bin(d_i - \sum_{k=1}^{j-1}r_{i,k}, p_{i,j}),
\end{align*}
where where and $j\leq i$ and $i=1,\ldots,T$.

\section{Inference}
As the main goal to generate inference of the number of death $\mv{d}$ is through the posterior distribution of number of deaths $\mv{d}$ given the observations $\mv{r}$.
In order to generate samples from this distribution we use a Markov Chain Monte Carlo method \citep{Brooks2011_handbook_markov}. In more detail we use a blocked Gibbs sampler, which generates samples in the following sequence:
\begin{itemize}
	\item  We sample $\mv{\alpha},\mv{\beta}, \mv{\alpha}^H,\mv{\beta}^H|\mv{d}, \mv{r}$ using the fact that one can integrate out $p$ in the model, and then  $\mv{d}|\mv{\alpha},\mv{\beta}, \mv{\alpha}^H,\mv{\beta}^H,\mv{r},\mv{\lambda}$  follows a Beta-Binomial distribution. Here to we use an adaptive MALA \citep{Atchade2006_adaptive_version} to sample from these parameters.
	\item  To sample $\mv{d}|\mv{\alpha},\mv{\beta}, \mv{\alpha}^H,\mv{\beta}^H,\mv{r},\mv{\lambda}$, that each death, $d_i$ is conditionally independent, and we just use a Metropolis Hastings random walk to sample each one.
	\item To sample $\mv{\lambda} | \mv{d},\sigma^2$ we again use an adaptive MALA.
	\item Finally We sample $\sigma^2|\mv{d}$,and  $p_0,\pi$ directly since this distribution is explicit, and $\phi$ using a MH-RW.
\end{itemize}

### 2020-06-11 output/paper/paper.tex
tex

\documentclass[a4paper,11pth,american]{article}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{babel}
\usepackage[protrusion=true,expansion=true,final,babel]{microtype}
\usepackage[babel]{csquotes}
\usepackage{babel}
\usepackage{natbib}
\usepackage{amsmath,amssymb,amsfonts,amsthm}
\usepackage{booktabs}
\usepackage{graphicx}
\DeclareGraphicsExtensions{.pdf,.png,.jpg,.eps}
\usepackage{array}
\usepackage{multirow}
\usepackage{tabularx}
\usepackage{hyperref}
\newcolumntype{L}{>{\raggedright\arraybackslash}X}
\newcommand{\mv}[1]{{\boldsymbol{#1}}}

@@

\begin{document}
    \maketitle
    \begin{abstract}
        The new corona virus disease --- COVID-2019 --- is rapidly spreading through the world. The availability of unbiased timely statistics of trends in disease events are a key to effective responses. But due to reporting delays, the most recently reported numbers are frequently underestimating of the total number of infections, hospitalizations and deaths creating an illusion of a downward trend. Here we describe a statistical methodology for predicting true daily quantities and their uncertainty, estimated using historical reporting delays. The methodology takes into account the observed distribution pattern of the lag. It is derived from the ``removal method'', a well-established estimation framework in the field of ecology.
	\end{abstract}
%%%%%%%%%%%%%%%%%%%
% MAIN TEXT SECTION
\section{Pandemic response demands timely data}
The new corona virus pandemic is affecting societies all around the world. As countries are challenged to control and fight back, they are in need of timely, unbiased, data for monitoring trends and making fast and well-informed decisions (\cite{No_author_2020_coronavirus_three}). Official statistics are usually reported with long delay after thorough verification, but in the midst of a deadly pandemic, real time data is of critical importance for policymakers (\cite{Jajosky2004_evaluation_reporting}). The latest data are often not finalized, but change as new information is reported. In fact, reporting delays make the most recent days have the least cases accounted for, producing a dangerous illusion of an always improving outlook.

@@

In this paper we describe a statistical methodology for nowcasting the epidemic statistics, such as hospitalizations or deaths, and their degrees of uncertainty, based on the daily reported event frequency and the observed distribution pattern of reporting delays. The prediction model is building on methods developed in ecology, referred to as the ``removal method'' (\cite{Pollock1991_review_papers}).

@@

\subsection{The current situation of COVID-19 in Sweden}
The Swedish Public Health Agency updates the COVID-19 statistics daily\footnote{The data is published on \url{https://www.folkhalsomyndigheten.se/smittskydd-beredskap/utbrott/aktuella-utbrott/covid-19/bekraftade-fall-i-sverige/}.}. During a press conference, they present updates on the number of deaths, admissions to hosiptals and intensive care, as well as case counts.

One of the reasons for following these indicators is to enable public health professionals and the public to observe the evolving patterns of the epidemic (\cite{Anderson2020_how_will}). In relation to policy, it is of specific interest to understand if the growth rates changes, which could indicate the need for a policy response. However, in each daily report only a proportion of the number of recent deaths is yet known, and this bias produces the illusion of a downward trend.

The death counts suffer from the longest reporting delay. In their daily press conference, the Swedish Public Health Agency warns for this by stopping the reported 7-day moving average trend line 10 days before the latest date. But not only are deaths often reported far further back than 10 days, a bar plot still shows the latest information, creating a sense of a downward trend. In fact, this might be the reason why the number of daily deaths have been underestimated repeatedly. At the peak, deaths were initially believed to level out at around 60 per day, but after all cases had been reported more than two weeks later, the actual number was close to 120 (\cite{Ohman2020_antalet_virusdoda}).

\section{The removal method}
We propose to use the removal method, developed in animal management (\cite{Pollock1991_review_papers}), to present an estimate of the actual frequencies at a given day and their uncertainty. The method has a long history dating back at least to the 1930s (\cite{Leslie1939_attempt_determine}). However, the first refined mathematical treatment of the method is credited to \cite{Moran1951_mathematical_theory}, more modern derivatives exits today (\cite{Matechou2016_open_models}). It is a commonly applied method today when analyzing age cohorts in fishery and wildlife management.

@@

A classic example where the method proposed to solve this problem has been used is in estimating statistics of trapping a closed population of animals (\cite{Pollock1991_review_papers}). Each day the trapped animals are collected, and kept, and if there is no immigration the number of trapped animals the following days will, on average, decline. This pattern of declining number of trapped animals allows one to draw inference of the underlying population size. Here we replace the animal population with the true number of deaths
on a given day. Instead of traps we have the new reports of COVID--19 events. As the number of new reported deaths for a given day declines, we can draw inference on how many actually died that day. If we assume that the reporting structure is constant over time we can after a while quickly get good estimate of the actual number.

@@

In Figure~\ref{fig:latest_prediction} we apply the model to the latest statistics from Sweden. The graph shows reported and predicted deaths (with uncertainty intervals) as bars, and a dashed line plots the 7-day (centered) moving average. A version without predictions is used in the Public Health Agency´s daily press briefings. As expected, the model provides estimates of actual deaths considerably above the reported number of deaths. Not how the model predicts additional deaths above the moving average line.

@@

Figure~\ref{fig:model_metrics} shows model performance compared to the benchmark for three difference performance metrics. All three graphs are based on predictions of reported deaths within 14 days, and show how performance increases as more data has been reported. Each data point is the average of all dates where predictions can be evaluated. SCRPS is a measure of accuracy that rewards precision, it is a proper scoring rule like the continuous probability rank score or the Brier score (see definition in Appendix) \cite{bolin2019scale}. The central plot shows the width of the prediction intervals, and the rightmost one the proportion of PI´s that cover the true value.

>>>>

\documentclass[a4paper,11pth]{article}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
%\usepackage{babel}
%\usepackage[protrusion=true,expansion=true,final,babel]{microtype}
%\usepackage[babel]{csquotes}
%\usepackage{babel}
\usepackage{natbib}
\usepackage{amsmath,amssymb,amsfonts,amsthm}
\usepackage{booktabs}
\usepackage{graphicx}
\DeclareGraphicsExtensions{.pdf,.png,.jpg,.eps}
\usepackage{array}
\usepackage{multirow}
\usepackage{tabularx}
\usepackage{hyperref}
\newcolumntype{L}{>{\raggedright\arraybackslash}X}
\newcommand{\mv}[1]{{\boldsymbol{#1}}}

@@

\begin{document}
    \maketitle
    \begin{abstract}
        The new corona virus disease --- COVID-2019 --- is rapidly spreading through the world. The availability of unbiased timely statistics of trends in disease events are a key to effective responses. But due to reporting delays, the most recently reported numbers are frequently underestimating of the total number of infections, hospitalizations and deaths creating an illusion of a downward trend. Here we describe a statistical methodology for predicting true daily quantities and their uncertainty, estimated using historical reporting delays. The methodology takes into account the observed distribution pattern of the lag. It is derived from the ``removal method'', a well-established estimation framework in the field of ecology.
	\end{abstract}
%%%%%%%%%%%%%%%%%%%
% MAIN TEXT SECTION
\section{Pandemic response demands timely data}
The new corona virus pandemic is affecting societies all around the world. As countries are challenged to control and fight back, they are in need of timely, unbiased, data for monitoring trends and making fast and well-informed decisions \citep{No_author_2020_coronavirus_three}. Official statistics are usually reported with long delay after thorough verification, but in the midst of a deadly pandemic, real time data is of critical importance for policymakers \citep{Jajosky2004_evaluation_reporting}. The latest data are often not finalized, but change as new information is reported. In fact, reporting delays make the most recent days have the least cases accounted for, producing a dangerous illusion of an always improving outlook.

@@

In this paper we describe a statistical methodology for nowcasting the epidemic statistics, such as hospitalizations or deaths, and their degrees of uncertainty, based on the daily reported event frequency and the observed distribution pattern of reporting delays. The prediction model is building on methods developed in ecology, referred to as the ``removal method'' \citep{Pollock1991_review_papers}.

@@

\subsection{The current situation of COVID-19 in Sweden}
The Swedish Public Health Agency updates the COVID-19 statistics daily\footnote{The data is published on \url{https://www.folkhalsomyndigheten.se/smittskydd-beredskap/utbrott/aktuella-utbrott/covid-19/bekraftade-fall-i-sverige/}.}. During a press conference, they present updates on the number of deaths, admissions to hospitals and intensive care, as well as case counts.

One of the reasons for following these indicators is to enable public health professionals and the public to observe the evolving patterns of the epidemic \citep{Anderson2020_how_will}. In relation to policy, it is of specific interest to understand if the growth rates changes, which could indicate the need for a policy response. However, in each daily report only a proportion of the number of recent deaths is yet known, and this bias produces the illusion of a downward trend.

The death counts suffer from the longest reporting delay. In their daily press conference, the Swedish Public Health Agency warns for this by stopping the reported 7-day moving average trend line 10 days before the latest date. But not only are deaths often reported far further back than 10 days, a bar plot still shows the latest information, creating a sense of a downward trend. In fact, this might be the reason why the number of daily deaths have been underestimated repeatedly. At the peak, deaths were initially believed to level out at around 60 per day, but after all cases had been reported more than two weeks later, the actual number was close to 120 \citep{Ohman2020_antalet_virusdoda}.

\section{The removal method}
We propose to use the removal method, developed in animal management \citep{Pollock1991_review_papers}, to present an estimate of the actual frequencies at a given day and their uncertainty. The method has a long history dating back at least to the 1930s \citep{Leslie1939_attempt_determine}. However, the first refined mathematical treatment of the method is credited to \citet{Moran1951_mathematical_theory}, more modern derivatives exits today \citep{Matechou2016_open_models}. It is a commonly applied method today when analyzing age cohorts in fishery and wildlife management.

@@

A classic example where the method proposed to solve this problem has been used is in estimating statistics of trapping a closed population of animals \citep{Pollock1991_review_papers}. Each day the trapped animals are collected, and kept, and if there is no immigration the number of trapped animals the following days will, on average, decline. This pattern of declining number of trapped animals allows one to draw inference of the underlying population size. Here we replace the animal population with the true number of deaths
on a given day. Instead of traps we have the new reports of COVID--19 events. As the number of new reported deaths for a given day declines, we can draw inference on how many actually died that day. If we assume that the reporting structure is constant over time we can after a while quickly get good estimate of the actual number.

@@

In Figure~\ref{fig:latest_prediction} we apply the model to the latest statistics from Sweden. The graph shows reported and predicted deaths (with uncertainty intervals) as bars, and a dashed line plots the 7-day (centered) moving average. A version without predictions is used in the Public Health Agency\'s daily press briefings. As expected, the model provides estimates of actual deaths considerably above the reported number of deaths. Not how the model predicts additional deaths above the moving average line.

@@

Figure~\ref{fig:model_metrics} shows model performance compared to the benchmark for three difference performance metrics. All three graphs are based on predictions of reported deaths within 14 days, and show how performance increases as more data has been reported. Each data point is the average of all dates where predictions can be evaluated. SCRPS is a measure of accuracy that rewards precision, it is a proper scoring rule like the continuous probability rank score or the Brier score (see definition in Appendix) \citep{bolin2019scale}. The central plot shows the width of the prediction intervals, and the rightmost one the proportion of PI\'s that cover the true value.
